---
title: "Distributional models for meta-analysis"
output: html_document
date: '2025-06-10'
---

# Required packages

```{r, warning=FALSE}
suppressMessages({
  library(here)
  library(dplyr)
  library(readr)
  library(tidyr) 
  library(tidyverse)
  library(purrr)
  library(metafor)
  library(brms)
  library(clubSandwich)
  }) 
# function
source(here("fun","fun.R"))
```

# Data

Load and process data.

```{r}
# load data
dat_raw <- readRDS(here("dat", "dat_medicine.RDS"))
# transform
dat <- lapply(dat_raw, function(df) {
  df %>% mutate(
    logOR = z2logOR(z),
    logOR.se = se_z2se_logOR(se_z = z.se, z = z)
  )
})

# add vi
dat <- lapply(dat, function(df) {
  df %>% mutate(
    logOR.var = logOR.se^2
  )
})

# saveRDS(dat, here("dat", "dat.RData"))
dat <- readRDS(here("dat", "dat.RData"))

```

# LS model

```{r}
# empty lists
res_list0 <- vector("list", length(dat))  
error_list0 <- vector("list", length(dat))  
warning_list0 <- vector("list", length(dat)) 

# fit
pb <- txtProgressBar(min = 0, max = length(dat), style = 3)
for (i in seq_along(dat)) {
  cat("\ndataset", i, "of", length(dat), "...\n") 
  
  tryCatch({
    res <- rma(logOR, logOR.var, scale = ~ 1, data = dat[[i]], 
               control = list(maxiter = 1000, stepadj = 0.5))
    
    res_list0[[i]] <- res
  },
  warning = function(w) {
    message("\n Error in dataset ", i, ": ", w$message)
    warning_list0[[i]] <- conditionMessage(w)  
    res_list0[[i]] <- NA  
  },
  error = function(e) {
    message("\n Warning in dataset ", i, ": ", e$message)
    error_list0[[i]] <- conditionMessage(e)  
    res_list0[[i]] <- NA 
  })
  
  setTxtProgressBar(pb, i)
}
close(pb)


#saveRDS(res_list0, here("dat", "res_list0.RData"))
res_list0 <- readRDS(here("dat", "res_list0.RData"))

# errors or warnings
#which(sapply(res_list0, is.null))
```

# First step

## Small-study effects1

```{r}
# empty lists
res_list1_PET <- vector("list", length(dat))  
error_list1 <- vector("list", length(dat))  
warning_list1 <- vector("list", length(dat)) 

# fit
pb <- txtProgressBar(min = 0, max = length(dat), style = 3)
for (i in seq_along(dat)) {
  cat("\ndataset", i, "of", length(dat), "...\n") 
  
  tryCatch({
    res <- rma(logOR, logOR.var, mods = ~ logOR.se, data = dat[[i]], 
               control = list(maxiter = 1000, stepadj = 0.5))
    
    res_list1_PET[[i]] <- res
  },
  warning = function(w) {
    message("\n Error in dataset ", i, ": ", w$message)
    warning_list1[[i]] <- conditionMessage(w)  
    res_list1_PET[[i]] <- NA  
  },
  error = function(e) {
    message("\n Warning in dataset ", i, ": ", e$message)
    error_list1[[i]] <- conditionMessage(e)  
    res_list1_PET[[i]] <- NA 
  })
  
  setTxtProgressBar(pb, i)
}
close(pb)


# saveRDS(res_list1_PET, here("dat", "res_list1_PET.RData"))
res_list1_PET <- readRDS(here("dat", "res_list1_PET.RData"))

# errors or warnings
#which(sapply(res_list1_PET, is.null))
```


## Small-study effects2

```{r}
# empty lists
res_list2_PET <- vector("list", length(dat))  
error_list2 <- vector("list", length(dat))  
warning_list2 <- vector("list", length(dat)) 

# fit
pb <- txtProgressBar(min = 0, max = length(dat), style = 3)
for (i in seq_along(dat)) {
  cat("\ndataset", i, "of", length(dat), "...\n") 
  
  tryCatch({
    res <- rma(logOR, logOR.var, mods = ~ logOR.se, scale = ~ logOR.se, data = dat[[i]], 
               control = list(maxiter = 1000, stepadj = 0.5, hessianCtrl=list(r=8)))
    
    res_list2_PET[[i]] <- res
  },
  warning = function(w) {
    message("\n Error in dataset ", i, ": ", w$message)
    warning_list2[[i]] <- conditionMessage(w)  
    res_list2_PET[[i]] <- NA  
  },
  error = function(e) {
    message("\n Warning in dataset ", i, ": ", e$message)
    error_list2[[i]] <- conditionMessage(e)  
    res_list2_PET[[i]] <- NA 
  })
  
  setTxtProgressBar(pb, i)
}
close(pb)


# save
# saveRDS(res_list2_PET, here("dat", "res_list2_PET.RData"))
res_list2_PET <- readRDS(here("dat", "res_list2_PET.RData"))

# errors or warnings
#which(sapply(res_list2_PET, is.null))
```

## Small-study effects3

```{r}
# empty lists
res_list3_PET <- vector("list", length(dat))  
error_list3 <- vector("list", length(dat))  
warning_list3 <- vector("list", length(dat)) 

# fit
pb <- txtProgressBar(min = 0, max = length(dat), style = 3)
for (i in seq_along(dat)) {
  cat("\ndataset", i, "of", length(dat), "...\n") 
  
  tryCatch({
    res <- rma(logOR, logOR.var, scale = ~ logOR.se, data = dat[[i]], 
               control = list(maxiter = 1000, stepadj = 0.5, hessianCtrl=list(r=8)))
    
    res_list3_PET[[i]] <- res
  },
  warning = function(w) {
    message("\n Error in dataset ", i, ": ", w$message)
    warning_list3[[i]] <- conditionMessage(w)  
    res_list3_PET[[i]] <- NA  
  },
  error = function(e) {
    message("\n Warning in dataset ", i, ": ", e$message)
    error_list3[[i]] <- conditionMessage(e)  
    res_list3_PET[[i]] <- NA 
  })
  
  setTxtProgressBar(pb, i)
}
close(pb)


# save
#saveRDS(res_list3_PET, here("dat", "res_list3_PET.RData"))
res_list3_PET <- readRDS(here("dat", "res_list3_PET.RData"))

# errors or warnings
#which(sapply(res_list3_PET, is.null))
```


## Small-study effects4

```{r}
# empty lists
res_list4_PET <- vector("list", length(dat))  
error_list4 <- vector("list", length(dat))  
warning_list4 <- vector("list", length(dat)) 

# fit
pb <- txtProgressBar(min = 0, max = length(dat), style = 3)
for (i in seq_along(dat)) {
  cat("\ndataset", i, "of", length(dat), "...\n") 
  
  tryCatch({
    res <- rma(logOR, logOR.var, scale = ~ logOR.var, data = dat[[i]], 
               control = list(maxiter = 1000, stepadj = 0.5, hessianCtrl=list(r=8)))
    
    res_list4_PET[[i]] <- res
  },
  warning = function(w) {
    message("\n Error in dataset ", i, ": ", w$message)
    warning_list4[[i]] <- conditionMessage(w)  
    res_list4_PET[[i]] <- NA  
  },
  error = function(e) {
    message("\n Warning in dataset ", i, ": ", e$message)
    error_list4[[i]] <- conditionMessage(e)  
    res_list4_PET[[i]] <- NA 
  })
  
  setTxtProgressBar(pb, i)
}
close(pb)


# save
#saveRDS(res_list4_PET, here("dat", "res_list4_PET.RData"))
res_list4_PET <- readRDS(here("dat", "res_list4_PET.RData"))

# errors or warnings
#which(sapply(res_list4_PET, is.null))
```



# Second step

## Small-study effects1

```{r}
# empty lists
res_list1_PEESE <- vector("list", length(dat))  
error_list1 <- vector("list", length(dat))  
warning_list1 <- vector("list", length(dat)) 

# fit
pb <- txtProgressBar(min = 0, max = length(dat), style = 3)
for (i in seq_along(dat)) {
  cat("\ndataset", i, "of", length(dat), "...\n") 
  
  tryCatch({
    res <- rma(logOR, logOR.var, mods = ~ logOR.var, data = dat[[i]], 
               control = list(maxiter = 1000, stepadj = 0.5))
    
    res_list1_PEESE[[i]] <- res
  },
  warning = function(w) {
    message("\n Error in dataset ", i, ": ", w$message)
    warning_list1[[i]] <- conditionMessage(w)  
    res_list1_PEESE[[i]] <- NA  
  },
  error = function(e) {
    message("\n Warning in dataset ", i, ": ", e$message)
    error_list1[[i]] <- conditionMessage(e)  
    res_list1_PEESE[[i]] <- NA 
  })
  
  setTxtProgressBar(pb, i)
}
close(pb)


#saveRDS(res_list1_PEESE, here("dat", "res_list1_PEESE.RData"))
res_list1_PEESE <- readRDS(here("dat", "res_list1_PEESE.RData"))

# errors or warnings
#which(sapply(res_list1_PEESE, is.null))
```


## Small-study effects2

```{r}
# empty lists
res_list2_PEESE <- vector("list", length(dat))  
error_list2 <- vector("list", length(dat))  
warning_list2 <- vector("list", length(dat)) 

# fit
pb <- txtProgressBar(min = 0, max = length(dat), style = 3)
for (i in seq_along(dat)) {
  cat("\ndataset", i, "of", length(dat), "...\n") 
  
  tryCatch({
    res <- rma(logOR, logOR.var, mods = ~ logOR.var, scale = ~ logOR.var, data = dat[[i]], 
               control = list(maxiter = 1000, stepadj = 0.5, hessianCtrl=list(r=8)))
    
    res_list2_PEESE[[i]] <- res
  },
  warning = function(w) {
    message("\n Error in dataset ", i, ": ", w$message)
    warning_list2[[i]] <- conditionMessage(w)  
    res_list2_PEESE[[i]] <- NA  
  },
  error = function(e) {
    message("\n Warning in dataset ", i, ": ", e$message)
    error_list2[[i]] <- conditionMessage(e)  
    res_list2_PEESE[[i]] <- NA 
  })
  
  setTxtProgressBar(pb, i)
}
close(pb)


# save
#saveRDS(res_list2_PEESE, here("dat", "res_list2_PEESE.RData"))
res_list2_PEESE <- readRDS(here("dat", "res_list2_PEESE.RData"))

# errors or warnings
#which(sapply(res_list2_PEESE, is.null))
```


## Small-study effects3

```{r}
# empty lists
res_list3_PEESE <- vector("list", length(dat))  
error_list3 <- vector("list", length(dat))  
warning_list3 <- vector("list", length(dat)) 

# fit
pb <- txtProgressBar(min = 0, max = length(dat), style = 3)
for (i in seq_along(dat)) {
  cat("\ndataset", i, "of", length(dat), "...\n") 
  
  tryCatch({
    res <- rma(logOR, logOR.var, scale = ~ logOR, data = dat[[i]], 
               control = list(maxiter = 1000, stepadj = 0.5, hessianCtrl=list(r=8)))
    
    res_list3_PEESE[[i]] <- res
  },
  warning = function(w) {
    message("\n Error in dataset ", i, ": ", w$message)
    warning_list3[[i]] <- conditionMessage(w)  
    res_list3_PEESE[[i]] <- NA  
  },
  error = function(e) {
    message("\n Warning in dataset ", i, ": ", e$message)
    error_list3[[i]] <- conditionMessage(e)  
    res_list3_PEESE[[i]] <- NA 
  })
  
  setTxtProgressBar(pb, i)
}
close(pb)


# save
#saveRDS(res_list3_PEESE, here("dat", "res_list3_PEESE.RData"))
res_list3_PEESE <- readRDS(here("dat", "res_list3_PEESE.RData"))

# errors or warnings
#which(sapply(res_list3_PEESE, is.null))
```


# Extraction

## Example

```{r}
# empty data frame
ests <- data.frame(
  estimate = rep(NA, length(res_list0)),
  se = rep(NA, length(res_list0)),
  pval = rep(NA, length(res_list0)),
  ci.lb = rep(NA, length(res_list0)),
  ci.ub = rep(NA, length(res_list0))
)


# extract
# normal model
for (i in seq_along(res_list0)) {
  model <- res_list[[i]]
  
  if (!is.null(model) && !is.na(model)) {
    ests$estimate[i] <- coef(model)[2]  
    ests$se[i] <- model$se[2]  
    ests$pval[i] <- model$pval[2] 
    ests$ci.lb[i] <- model$ci.lb[2]  
    ests$ci.ub[i] <- model$ci.ub[2] 
  }
}

# intercept-only ls model
res <- rma(logOR, logOR.var, data = dat[[11]], 
               control = list(maxiter = 1000, stepadj = 0.5))
res.ls <- rma(logOR, logOR.var, scale = ~ 1, data = dat[[11]], 
               control = list(maxiter = 1000, stepadj = 0.5))


for (i in seq_along(res_list0)) {
  model <- res_list0[[i]]
  
  if (!is.null(model) && !is.na(model)) {
    ests$estimate[i] <- coef(model)$alpha[1]  
    ests$se[i] <- model$se.alpha[1] 
    ests$pval[i] <- model$pval.alpha[1] 
    ests$ci.lb[i] <- model$ci.lb.alpha[1]  
    ests$ci.ub[i] <- model$ci.ub.alpha[1] 
  }
}


# regression ls model
for (i in seq_along(res_list0)) {
  model <- res_list0[[i]]
  
  if (!is.null(model) && !is.na(model)) {
    ests$estimate[i] <- coef(model)$alpha[2][1]  
    ests$se[i] <- model$se.alpha[2] 
    ests$pval[i] <- model$pval.alpha[2] 
    ests$ci.lb[i] <- model$ci.lb.alpha[2]  
    ests$ci.ub[i] <- model$ci.ub.alpha[2] 
  }
}


```


## Publication bias

```{r}
# small study effect
# empty data frame
ests_pb1 <- data.frame(
  k = rep(NA, length(res_list1_PET)),
  estimate = rep(NA, length(res_list1_PET)),
  se = rep(NA, length(res_list1_PET)),
  pval = rep(NA, length(res_list1_PET)),
  ci.lb = rep(NA, length(res_list1_PET)),
  ci.ub = rep(NA, length(res_list1_PET))
)

# extract
for (i in seq_along(res_list1_PET)) {
  model <- res_list1_PET[[i]]
  
  if (!is.null(model) && !is.na(model)) {
    ests_pb1$k[i] <- model$k
    ests_pb1$estimate[i] <- coef(model)[2]  
    ests_pb1$se[i] <- model$se[2]  
    ests_pb1$pval[i] <- model$pval[2] 
    ests_pb1$ci.lb[i] <- model$ci.lb[2]  
    ests_pb1$ci.ub[i] <- model$ci.ub[2] 
  }
}

#saveRDS(ests_pb1, here("dat", "ests_pb1.RData"))
ests_pb1 <- readRDS(here("dat", "ests_pb1.RData"))
# add label
ests_pb1$MA.id <- names(dat)
length(which(ests_pb1$pval < 0.05))

# remove NA and extreme estimate
ests_pb1 <- ests_pb1[!is.na(ests_pb1$estimate),]
ests_pb1 <- ests_pb1 %>% filter(estimate < 50 & estimate > -50)

length(which(ests_pb1$pval < 0.05)) / nrow(ests_pb1)

# small study effect2
# empty data frame
ests_pb2 <- data.frame(
  k = rep(NA, length(res_list2_PET)),
  estimate = rep(NA, length(res_list2_PET)),
  se = rep(NA, length(res_list2_PET)),
  pval = rep(NA, length(res_list2_PET)),
  ci.lb = rep(NA, length(res_list2_PET)),
  ci.ub = rep(NA, length(res_list2_PET)),
  estimate_alpha = rep(NA, length(res_list2_PET)),
  se_alpha = rep(NA, length(res_list2_PET)),
  pval_alpha = rep(NA, length(res_list2_PET)),
  ci.lb_alpha = rep(NA, length(res_list2_PET)),
  ci.ub_alpha = rep(NA, length(res_list2_PET))
)

# extract
for (i in seq_along(res_list2_PET)) {
  model <- res_list2_PET[[i]]
  
  if (!is.null(model) && !is.na(model)) {
    ests_pb2$k[i] <- model$k
    ests_pb2$estimate[i] <- coef(model)$beta[2]  
    ests_pb2$se[i] <- model$se[2]  
    ests_pb2$pval[i] <- model$pval[2] 
    ests_pb2$ci.lb[i] <- model$ci.lb[2]  
    ests_pb2$ci.ub[i] <- model$ci.ub[2]
    
    ests_pb2$estimate_alpha[i] <- coef(model)$alpha[2]  
    ests_pb2$se_alpha[i] <- model$se.alpha[2]  
    ests_pb2$pval_alpha[i] <- model$pval.alpha[2] 
    ests_pb2$ci.lb_alpha[i] <- model$ci.lb.alpha[2]  
    ests_pb2$ci.ub_alpha[i] <- model$ci.ub.alpha[2]
  }
}

#saveRDS(ests_pb2, here("dat", "ests_pb2.RData"))
ests_pb2 <- readRDS(here("dat", "ests_pb2.RData"))
# add label
ests_pb2$MA.id <- names(dat)
length(which(ests_pb2$pval < 0.05))
length(which(ests_pb2$pval_alpha < 0.05))


# remove NA and extreme estimate
ests_pb2 <- ests_pb2[!is.na(ests_pb2$estimate),]
ests_pb2 <- ests_pb2 %>% filter(estimate < 50 & estimate > -50)


length(which(ests_pb2$pval < 0.05)) / nrow(ests_pb2)
length(which(ests_pb2$pval_alpha < 0.05)) / nrow(ests_pb2)

# small study effect3
# empty data frame
ests_pb3 <- data.frame(
  k = rep(NA, length(res_list3_PET)),
  estimate_alpha = rep(NA, length(res_list3_PET)),
  se_alpha = rep(NA, length(res_list3_PET)),
  pval_alpha = rep(NA, length(res_list3_PET)),
  ci.lb_alpha = rep(NA, length(res_list3_PET)),
  ci.ub_alpha = rep(NA, length(res_list3_PET))
)

# extract
for (i in seq_along(res_list3_PET)) {
  model <- res_list3_PET[[i]]
  
  if (!is.null(model) && !is.na(model)) {
    ests_pb3$k[i] <- model$k
    ests_pb3$estimate_alpha[i] <- coef(model)$alpha[2]  
    ests_pb3$se_alpha[i] <- model$se.alpha[2]  
    ests_pb3$pval_alpha[i] <- model$pval.alpha[2] 
    ests_pb3$ci.lb_alpha[i] <- model$ci.lb.alpha[2]  
    ests_pb3$ci.ub_alpha[i] <- model$ci.ub.alpha[2]
  }
}

#saveRDS(ests_pb3, here("dat", "ests_pb3.RData"))
ests_pb3 <- readRDS(here("dat", "ests_pb3.RData"))

# add label
ests_pb3$MA.id <- names(dat)
length(which(ests_pb3$pval_alpha < 0.05))


# small study effect4
# empty data frame
ests_pb4 <- data.frame(
  k = rep(NA, length(res_list4_PET)),
  estimate_alpha = rep(NA, length(res_list4_PET)),
  se_alpha = rep(NA, length(res_list4_PET)),
  pval_alpha = rep(NA, length(res_list4_PET)),
  ci.lb_alpha = rep(NA, length(res_list4_PET)),
  ci.ub_alpha = rep(NA, length(res_list4_PET))
)

# extract
for (i in seq_along(res_list4_PET)) {
  model <- res_list4_PET[[i]]
  
  if (!is.null(model) && !is.na(model)) {
    ests_pb4$k[i] <- model$k
    ests_pb4$estimate_alpha[i] <- coef(model)$alpha[2]  
    ests_pb4$se_alpha[i] <- model$se.alpha[2]  
    ests_pb4$pval_alpha[i] <- model$pval.alpha[2] 
    ests_pb4$ci.lb_alpha[i] <- model$ci.lb.alpha[2]  
    ests_pb4$ci.ub_alpha[i] <- model$ci.ub.alpha[2]
  }
}

#saveRDS(ests_pb4, here("dat", "ests_pb4.RData"))
ests_pb4 <- readRDS(here("dat", "ests_pb4.RData"))

# add label
ests_pb3$MA.id <- names(dat)
length(which(ests_pb3$pval_alpha < 0.05))


```


## True effect approximation

```{r}
est.petpeese_SMD.zr <- do.call(rbind, lapply(seq_along(adjustPET_SMD.zr_RVE), function(i) {
  # NULL models for PET and PEESE
  PET_model <- adjustPET_SMD.zr_RVE[[i]]
  PEESE_model <- adjustPEESE_SMD.zr_RVE[[i]]
  
  if (is.null(PET_model) || is.null(PEESE_model)) {
    # assign NA if any model is null
    return(data.frame(est = NA, se = NA, p = NA, source = NA))
  }
  
  tryCatch({
    # evaluate PET p-value
    PET_pval <- PET_model$pval[1]
    
    if (PET_pval > 0.1) {
      # extract from PET if test is not rejected
      return(data.frame(
        est = PET_model$b[1],
        se  = PET_model$se[1],
        p   = PET_pval,
        source = "PET"
      ))
    } else if (PET_pval < 0.1) {
      # extract from PEESE if PET is rejected
      return(data.frame(
        est = PEESE_model$b[1],
        se  = PEESE_model$se[1],
        p   = PEESE_model$pval[1],
        source = "PEESE"
      ))
    } else {
      # assign NA if conditions are ambiguous
      return(data.frame(est = NA, se = NA, p = NA, source = NA))
    }
  }, error = function(e) {
    # errors during extraction
    return(data.frame(est = NA, se = NA, p = NA, source = NA))
  })
}))

#saveRDS(est.petpeese_SMD.zr, here("Dat", "est.petpeese_SMD.zr.RData"))
est.petpeese_SMD.zr <- readRDS(here("Dat", "est.petpeese_SMD.zr.RData"))
```


# Software and package versions

```{r}
knitr::kable(installed.packages()[names(sessionInfo()$otherPkgs), "Version"], "pipe")
```

|package      |version    |
|:------------|:----------|
|brms         |2.22.0     |
|Rcpp         |1.0.13     |
|clubSandwich |0.5.11     |
|metafor      |4.7-53     |
|numDeriv     |2016.8-1.1 |
|metadat      |1.2-0      |
|Matrix       |1.5-3      |
|forcats      |0.5.2      |
|stringr      |1.5.0      |
|purrr        |0.3.4      |
|tibble       |3.1.8      |
|ggplot2      |3.4.4      |
|tidyverse    |1.3.1      |
|tidyr        |1.2.1      |
|readr        |2.1.2      |
|dplyr        |1.0.10     |
|here         |1.0.1      |